---
title: "Week 3: Graphics and Visualization"
subtitle: "MATH-517 Statistical Computation and Visualization"
author: "Tomas Masak"
date: "`r format(Sys.time(), '%b %d, %Y')`"
#date: "September 21st 2021"
output: beamer_presentation
classoption: "presentation"
theme: "Madrid"
colortheme: "seahorse"
footer: "Copyright (c) 2022, EPFL"
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(bookdown)
library(tidyverse)
library(data.table)
theme_set(theme(text=element_text(size=20)))
```

## Graphics

* presentation
  - result communication
  - decision making
* data insight
  - large data
  - detect patterns
  - find strange observations
* code debugging
  - input, output (even the code itself) is data

## Graphics

> "The simple graph has brought more information to the data analyst’s mind than any other device." -- John W. Tukey

&nbsp;

> "The greatest value of a picture is when it forces us to notice what we never expected to see." -- John W. Tukey

&nbsp;

\begin{exampleblock}{}
\centering \textbf{One can think of graphics (and also models, for that matter) as of a low-dimensional representation for data.}
\end{exampleblock}

## Anscombe's Quartet

Four data sets with

* one response variable $y$
* one regressor $x$

&nbsp;

```{r}
data(anscombe) # attach data
lm1 <- lm(y1 ~ x1, data = anscombe)
lm2 <- lm(y2 ~ x2, data = anscombe)
lm3 <- lm(y3 ~ x3, data = anscombe)
lm4 <- lm(y4 ~ x4, data = anscombe)
lm1 <- c(coef(lm1),summary(lm1)$r.squared)
lm2 <- c(coef(lm2),summary(lm2)$r.squared)
lm3 <- c(coef(lm3),summary(lm3)$r.squared)
lm4 <- c(coef(lm4),summary(lm4)$r.squared)
names(lm1) <- c("(Intercept)","x","R-squared")
rbind(lm1,lm2,lm3,lm4)
```

## Anscombe's Quartet

```{r, fig.show='hold', out.width='40%',message=F, warning=F, fig.align='center'}
library(tidyverse)
ggplot(data=anscombe, mapping=aes(x=x1,y=y1)) +
  geom_point(size=3) +
  stat_smooth(method='lm') +
  theme(text=element_text(size=20), )
ggplot(data=anscombe, mapping=aes(x=x2,y=y2)) +
  geom_point(size=3) +
  stat_smooth(method='lm') +
  theme(text=element_text(size=20), )
ggplot(data=anscombe, mapping=aes(x=x3,y=y3)) +
  geom_point(size=3) +
  stat_smooth(method='lm') +
  theme(text=element_text(size=20), )
ggplot(data=anscombe, mapping=aes(x=x4,y=y4)) +
  geom_point(size=3) +
  stat_smooth(method='lm') +
  theme(text=element_text(size=20), )
```

## Human Height

\begin{center}
\includegraphics[width=0.9\textwidth]{../Plots/human_height.jpg}
\end{center}

## Plot Dimension

```{r, out.width="45%",fig.show='hold'}
library(lattice)
x <- 1:30/31
y1 <- cos(2*pi*x)
y2 <- x
y <- outer(y1,y2)
persp(x=x,y=x,z=y,theta=180)
levelplot(y)

y2 <- sin(4*pi*x)
y <- outer(y1,y2)
persp(x=x,y=x,z=y,theta=180)
levelplot(y)
```

## Scatterplot Extras

The following features (`ggplot`'s arguments) for points (and similarly for lines)

* `color`
* `shape`
* `size`
* `alpha` (opacity/transparency)

can be used

* to include additional information (or dimensions, i.e. to include additional variables) in a scatterplot
* to combat overplotting
* or simply to make the plot nicer (i.e. used subjectively)

## Available Shapes

\begin{center}
\includegraphics[width=0.5\textwidth]{../Plots/pch.png}
\end{center}

* `pch` is the base R argument, for `ggplot` one instead passes values into `scale_shape_manual()`
* all shapes have attribute `color`, only some `fill`

## Scatterplot: plain

\footnotesize
```{r, echo=T, out.width="60%",fig.align="center"}
library(data.table)
set.seed(517)
flights <- fread("https://raw.githubusercontent.com/Rdatatable/data.table/master/vignettes/flights14.csv") %>%
  slice_sample(n=200) 
ggplot(data = flights,
       mapping = aes(x = arr_delay, y = distance)) +
  geom_point()
```

## Scatterplot: shape + color

\footnotesize
```{r, echo=T, out.width="70%",fig.align="center"}
ggplot(data = flights, mapping = aes(x = arr_delay, y = distance,
                                     shape = origin, fill=origin)) +
  geom_point(size=2) + 
  scale_fill_manual(values = c("cyan","red","green")) +
  scale_shape_manual(values = c(21,22,24))
```

## Scatterplot: shape $\times$ color

\footnotesize
```{r, echo=T, out.width="70%",fig.align="center"}
ggplot(data = flights, mapping = aes(x = arr_delay, y = distance, 
                                     shape = origin, color=carrier)) +
  geom_point(size=2) + scale_shape_manual(values = c(15,16,17))
```

## Overplotting

\footnotesize
```{r, echo=T, out.width="60%",fig.align="center"}
set.seed(517)
flights <- fread("https://raw.githubusercontent.com/Rdatatable/data.table/master/vignettes/flights14.csv") %>%
  slice_sample(n=5000) 
ggplot(data = flights, mapping = aes(x = arr_delay, y = distance,
                                     shape = origin, fill=origin)) +
  geom_point(size=2) + 
  scale_fill_manual(values = c("cyan","red","green")) +
  scale_shape_manual(values = c(21,22,24))
```

## Overplotting Remedy 1/2

\footnotesize
```{r, echo=T, fig.dim=c(18,6), out.width="100%",fig.align="center"}
ggplot(data = flights, mapping = aes(x = arr_delay, y = distance,
                                     shape = origin, fill=origin)) +
  geom_point() + 
  scale_fill_manual(values = c("cyan","red","green")) +
  scale_shape_manual(values = c(21,22,24)) +
  facet_wrap(~origin)
```

## Overplotting Remedy

\footnotesize
```{r, echo=T, fig.dim=c(12,4), out.width="100%",fig.align="center"}
ggplot(data = flights, mapping = aes(x = arr_delay, y = distance)) +
  stat_binhex() + 
  scale_fill_gradient(low = "lightblue", high = "red") +
  facet_wrap(~origin)
```

## Overplotting Remedies

* reducing point size
* split a single plot into multiple (done above)
* jittering
* opacity/transparency
* binning (done above via `stat_binhex()`)
* tiles
* subsampling (done above above)

Many of these options clash with clarity and space requirements or with each other (e.g. transparency distorts colors).

## Visualizing Variance

\centering
\includegraphics[width=0.9\textwidth]{../Plots/variance_plotting/line_ok.png}

## Visualizing Variance (Bad)

\centering
\includegraphics[width=0.9\textwidth]{../Plots/variance_plotting/line_bad.png}

* the grey lines should be much thinner and colored so we can relate them to the colored means (and x-axis added)

## Visualizing Variance

\centering
\includegraphics[width=0.9\textwidth]{../Plots/variance_plotting/bands.png}

## Visualizing Variance

\centering
\includegraphics[width=0.9\textwidth]{../Plots/variance_plotting/boxplots.png}
  
## Spatial Data

Complicated due to different

* mathematical representations
* data structures
* data sources
* data processing packages
  - projections (sphere? plane?)
* visualization packages

(Not so) short course about visualizing spatial data [here](https://github.com/Robinlovelace/Creating-maps-in-R) (only if interested).

## Mercator Projection

\includegraphics[width=\textwidth]{../Plots/mercator.png}

\hfill \textcolor{gray}{source: wiki}

Two sources of distortion (earth is not a sphere and the projection itself), the second one visualized [here](https://en.wikipedia.org/wiki/Mercator_projection#/media/File:Worlds_animate.gif).

## Russia's True Size

\centering
\includegraphics[width=0.75\textwidth]{../Plots/russia_true_size.jpg}

## Geospatial Heatmap

\footnotesize
```{r, echo=T, eval=F}
library(ggswissmaps)
counties <- shp_df[[4]] # resolution on the level of ZIP codes

avg_age <- read.csv("https://raw.githubusercontent.com/grssnbchr/thematic-maps-ggplot2/master/input/avg_age_15.csv")
names(avg_age)[1] <- "id"
names(avg_age)[3] <- "age"
counties <- counties %>% mutate(id = as.numeric(id))
newdat <- inner_join(counties, avg_age, by="id")
library(viridis)
ggplot(data=newdat) +
  geom_polygon(aes(x=long, y=lat, group=group, fill=age)) +
  geom_path(aes(x = long, y = lat, group = group),
            color = "white", size = 0.1) +
  scale_fill_viridis(option = "magma", direction = -1) +
  theme_void()
```

* here counties is a data frame, longitude and latitude specify border points and Swiss counties are polygons (convex hulls of the border points)
* `group` specifies which border points belong to which county
* try `plot(counties$long,counties$lat, type="l")`

## Geospatial Heatmap

```{r, out.width="80%", fig.align="center"}
library(ggswissmaps)
counties <- shp_df[[4]] # resolution on the level of ZIP codes
# now read average age data for ZIP regions
avg_age <- read.csv("https://raw.githubusercontent.com/grssnbchr/thematic-maps-ggplot2/master/input/avg_age_15.csv")
names(avg_age)[1] <- "id"
names(avg_age)[3] <- "age"
counties <- counties %>% mutate(id = as.numeric(id))
newdat <- inner_join(counties, avg_age, by="id") # merge two data frames
library(viridis) # just for a nicer color scale
ggplot(data=newdat) +
  geom_polygon(aes(x=long, y=lat, group=group, fill=age)) +
  geom_path(aes(x = long, y = lat, group = group),
            color = "white", size = 0.1) +
  scale_fill_viridis(option = "magma", direction = -1) +
  theme_void()
```

\hfill inspired by [this blogpost](https://timogrossenbacher.ch/2016/12/beautiful-thematic-maps-with-ggplot2-only/)

## Good Visualization Practices

* provide context (in text **and** in caption)
* seek simplicity, clarity, etc.
* gray scale often preferable
  - color-blindness (friendly palettes, e.g. [Coolors](https://coolors.co/333333-666a86-95b8d1-e8ddb5-edafb8))
* axes (scale, gaps, etc.)
  - text of appropriate size
* publication-specific conditions
* be artistic!
  - sometimes bend the rules (responsibly and justifiably)
  
Find inspiration in [The R Graph Gallery](https://r-graph-gallery.com/).

Beware when exporting graphics.

# Bad Visualization Practices

## Reverted Axis

\centering
\includegraphics[width=0.9\textwidth]{../Plots/bad_plotting/1.png}

## False Venn's Diagrams

\centering
\includegraphics[width=0.9\textwidth]{../Plots/bad_plotting/2.png}

## Missing Baseline

\centering
\includegraphics[width=0.6\textwidth]{../Plots/bad_plotting/3.jpg}

## Axis Starting at Zero

\centering
\includegraphics[width=0.9\textwidth]{../Plots/bad_plotting/4.png}

## Better

\centering
\includegraphics[width=0.9\textwidth]{../Plots/bad_plotting/5.png}

## Combined Effects

\centering
\includegraphics[width=0.9\textwidth]{../Plots/bad_plotting/6.png}

## ...plus just Cheating

\centering
\includegraphics[width=0.9\textwidth]{../Plots/bad_plotting/7.jpg}

## Double Axes

\centering
\includegraphics[width=0.9\textwidth]{../Plots/bad_plotting/8.png}

* This is actually quite good, but double axes are usually problematic.

## Tweaking Axis

\centering
\includegraphics[width=0.9\textwidth]{../Plots/bad_plotting/9.png}

## Binning

\centering
\includegraphics[width=0.9\textwidth]{../Plots/bad_plotting/10.png}

* Binning itself is not bad, but it is basically a variable (axis) transformation, so beware.

## Different Kinds of Binning

\centering
\includegraphics[width=0.9\textwidth]{../Plots/bad_plotting/11.png}

## Total

\centering
\includegraphics[width=0.9\textwidth]{../Plots/bad_plotting/12.png}

## Relative

\centering
\includegraphics[width=0.9\textwidth]{../Plots/bad_plotting/13.png}

## Useless 3D

\centering
\includegraphics[width=0.9\textwidth]{../Plots/bad_plotting/14.png}

## Better 2D

\centering
\includegraphics[width=0.9\textwidth]{../Plots/bad_plotting/15.png}

## Total vs. Relative Again

\centering
\includegraphics[width=0.7\textwidth]{../Plots/bad_plotting/16.png}

## Missing Axis & Misguidance

\centering
\includegraphics[width=0.9\textwidth]{../Plots/bad_plotting/17.png}

## One more Scatterplot Example

* voting records of the members of the Czech parliament in 2015
  - $N=200$ members of the parliament
  - certain number of votings during that period ($x_{nj} \in \{ -1,0,1 \}$ depending on whether $n$-th member voted for the $j$-th voting no/abstain/yes), but **PCA** applied and only first 2 components kept
* additional information about party affiliation of the members
  - `party` affiliation (7 parties in total)
  - government `coalition` affiliation (T/F)
  - left-wing/right-wing scale `LRscale` (on the party level)

```{r}
load("../data/cz_parliament_2015.RData")
cz$LRscale <- as.factor(cz$LRscale)
str(cz)
```

## One more Scatterplot Example

\footnotesize
```{r, echo=T, eval=F}
ggplot(data = cz, mapping = aes(x = PC1, y = PC2, color=party,
                                shape=coalition, size=LRscale)) +
  geom_point() +
  scale_color_manual(values=c("skyblue","orange","yellow",
                              "red","blue","blueviolet","green3")) +
  scale_shape_manual(values=c(16,17))
```

## One more Scatterplot Example

\footnotesize
```{r, out.width="90%",fig.align="center"}
load("../data/cz_parliament_2015.RData")
cz$LRscale <- as.factor(cz$LRscale)
ggplot(data = cz, mapping = aes(x = PC1, y = PC2, color=party, shape=coalition, size=LRscale)) +
  geom_point() +
  scale_color_manual(values=c("skyblue","orange","yellow","red","blue","blueviolet","green3")) +
  scale_shape_manual(values = c(16,17))
```

## Asignment

**Small project [20%]**. Deadline on Week 6.

The goal of this project is *data exploration*. Find an interesting (in the sense it interests you!) data set and

* explore the data
* describe the data
* visualize the data
* lay out some questions about the data based on your exploration
* use more detailed visualization techniques to hint answers

The first step should be done individually. Then you can form groups of 2-3 and pick up the most interesting data set and do the rest. See [Course organization](../Manuals/01_Organization.html) for more details.

## Exercise

1. Some data repositories listed on following slides will be explored.

2. Consider the Simpsons IMDB ratings plot above. Choose say 3 of your favorite TV shows, download data from the [IMDB database](https://www.imdb.com/interfaces/) a produce a plot similar to the Simpsons plot above (i.e. variances captured with boxplots) using `ggplot()` depicting all 3 TV shows.

3. Find and obtain spatial boundary files of administrative regions of your home country (or use some other country, if problematic). Produce a geospatial heatmap of the country, using some external variable for the color fill argument (e.g. specific election results or election participation, mortality rates, etc.).

4. Choose a scatterplot above (say the one on slide 10) and try to reproduce it using base R functions `plot()` instead of `ggplot()` and add legend manually using `legend()`.

## Exercise: Some Links to Open Data

[fivethirtyeight](https://data.fivethirtyeight.com/): article data of Nate Silver's data journalism platform freely available (see also R package [fivethirtyeight](https://cran.r-project.org/web/packages/fivethirtyeight/index.html))

[data-is-plural](https://www.data-is-plural.com/): weekly newsletter of datasets by Jeremy Singer-Vine

[re3data](https://www.re3data.org/): Registry of research data repositories

[openml datasets](https://www.openml.org/): many uniformly formatted datasets for training machine learning models -- however, not always good descriptions available

[Worldbank Datacatalog](https://datacatalog.worldbank.org/): the World Bank data catalogue

[UK Data Service](https://beta.ukdataservice.ac.uk/datacatalogue/studies#!?Search=&Rows=10&Sort=0&DateFrom=1753&DateTo=2018&AccessFacet=Open&Page=1): UK’s largest collection of social, economic and population data resources (filter for open data) or also [data.gov.uk](https://www.data.gov.uk/)

[ICPSR](https://www.icpsr.umich.edu): unit within the Institute for Social Research at the University of Michigan, social and behavioral research. In particular including [replication datasets](https://www.openicpsr.org/openicpsr/) for published studies.

[govdata](https://www.govdata.de/): Open Government - German administrative data freely accessible

## Exercise: Some more

[gapminder](https://www.gapminder.org/): "an independent educational non-proﬁt ﬁghting global misconceptions"; collection and vizualisation of datasets concerning gobal developement

[nature.com](https://www.nature.com/sdata/): peer-reviewed, open-access journal for descriptions of datasets (broad range of natural science disciplines)

[NIH (National Institute of Health) Data Sharing Repositories](https://www.nlm.nih.gov/NIHbmic/domain_specific_repositories.html): overview on different thematically sorted medical databases

[UCI Machine Learning Repository](https://archive.ics.uci.edu/ml/datasets.php) or the new [beta version](https://archive-beta.ics.uci.edu/): containing various datasets -- however, sometimes with a little few description

[data.bris Research Data Repository](https://data.bris.ac.uk/data/): Data repository of the University of Bristol

[bellingcat TikTok Hashtag analysis tool](https://www.bellingcat.com/resources/how-tos/2022/05/11/this-new-tool-lets-you-analyse-tiktok-hashtags/): Didn't try, don't know how easy.

_... no systematic selection. Much more out there_
